The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-02-19 paper.tex
Theorem provers/synthesis

\subparagraph{Formal model.}
The options generated by The Gamma let the user select both the next operation and the parameters
of the previously selected operation. The available operations and parameters are generated
based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
contains the current schema $S$ and the operations applied so far. In the following, we write
$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
\[
\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
\]
The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
expects further parameters or whether it is fully-specified. In the first case, the recommendation
engine generates possible additional parameter values $p_', p_'', \ldots$ based on the schema $S$,
the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
The $\choices$ operation then generates options that add the additional parameter. We generated the
identifiers $\iota_',\iota_'',\ldots$ based on the state and the parameter value, such as
\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
\[
\begin{array}{l}
\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
\qquad\;\;    \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots~\}
\end{array}
\]
If the last operation takes no further parameters, the system produces a choice
of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
operation then returns options that add the additional operation:
\[
\begin{array}{l}
\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
\qquad \;\; \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots~\}
\end{array}
\]
Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
host language lets us compose functions using $f\circ g$, we can write:
\[
\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
\]
The recommendation engine behind The Gamma provides a domain-specific logic for generating
possible operations and their parameters based on the current schema of the data. As the above
definition shows, this underlying engine can be easily exposed through the common
choose-your-own-adventure interface.

@@

\noindent
For a given set of constraints $\boldsymbol{c}$, the $\select$ operation looks
for $e\in E_{\boldsymbol{c}}$ with the largest score:
\[
\select_X(\boldsymbol{c}) = \argmax_{e \in E_{\boldsymbol{c}}} Q_{\boldsymbol{c}}(X, e)
\]
The actual implementation of the optimization uses various machine learning techniques to
find the optimal expression. In case of datadiff, $X$ is a pair of datasets $X_1, X_2$ to be
reconciled. The AI assistant uses the Hungarian algorithm \cite{sutton-2018-datadiff} to construct
a matching of columns from $X_1$ and $X_2$. The generated expression is a sequence of patches
that can be applied to $X_2$ in order to reconcile its structure with the sturcture of $X_1$.
The constraints specified by the user restrict the space of possible column matchings and so they
affect $E_\boldsymbol{c}$. The scoring function $Q_\boldsymbol{c}$ is independent of
the constraints and computes a sum of distances between the statistical distributions of the
columns from $X_1$ and a patched version of $X_2$.

@@

\subsection{Interactive theorem proving}
~

The internal mode can also be used when
working with theorem provers.

In Coq, where proofs are represented as series of tactics,

However, this would also be case in a
completion engine for the Coq proof system that would offer a choice of tactics to apply, because
Coq proofs are represented

@@

@@

>>>>

\subparagraph{Formal model.}
The options generated by The Gamma let the user select both the next operation and the parameters
of the previously selected operation. The available operations and parameters are generated
based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
contains the current schema $S$ and the operations applied so far. In the following, we write
$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
\[
\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
\]
The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
expects further parameters or whether it is fully-specified. In the first case, the recommendation
engine generates possible additional parameter values $p', p'', \ldots$ based on the schema $S$,
the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
The $\choices$ operation then generates options that add the additional parameter. We generated the
identifiers $\iota',\iota'',\ldots$ based on the state and the parameter value, such as
\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
\[
\begin{array}{l}
\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
\qquad\;\;    \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots~\}
\end{array}
\]
If the last operation takes no further parameters, the system produces a choice
of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
operation then returns options that add the additional operation:
\[
\begin{array}{l}
\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
\qquad \;\; \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots~\}
\end{array}
\]
Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
host language lets us compose functions using $f\circ g$, we can write:
\[
\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
\]
The recommendation engine behind The Gamma provides a domain-specific logic for generating
possible operations and their parameters based on the current schema of the data. As the above
definition shows, this underlying engine can be easily exposed through the common
choose-your-own-adventure interface.

@@

\noindent
For a given set of constraints $\boldsymbol{c}$, the $\select$ operation looks
for $e\in E_{\boldsymbol{c}}$ with the largest score:
\[
\select_X(\boldsymbol{c}) = \argmax_{e \in E_{\boldsymbol{c}}} Q_{\boldsymbol{c}}(X, e)
\]
The actual implementation of the optimization uses various machine learning techniques to
find the optimal expression. In case of datadiff, $X$ is a pair of datasets $X_1, X_2$ to be
reconciled. The AI assistant uses the Hungarian algorithm \cite{sutton-2018-datadiff} to construct
a matching of columns from $X_1$ and $X_2$. The generated expression is a sequence of patches
that can be applied to $X_2$ in order to reconcile its structure with the sturcture of $X_1$.
The constraints specified by the user restrict the space of possible column matchings and so they
affect $E_{\boldsymbol{c}}$. The scoring function $Q_{\boldsymbol{c}}$ is independent of
the constraints and computes a sum of distances between the statistical distributions of the
columns from $X_1$ and a patched version of $X_2$.

@@

\subsection{Theorem proving}

In the previous three sections, we showed how existing formally well-documented systems fit the
choose-your-own-adventure interaction pattern. Although interactive theorem provers and editors
for dependently typed languages implement similar kinds of interactions, there is no well-documented
system that exactly fits the pattern. The closest example is perhaps the recently envisioned
mixed-mode interaction theorem prover \cite{verter-2024-mixed}. Rather than reframing the
implementation of an existing system, this section thus outlines a possible implementation.

\subparagraph{Theorem provers.}
There are two approaches to interacting with an interactive theorem prover. In Coq, the user
writes a sequence of tactics that transform proof goals. In Agda or Idris, the user writes a
term, or program, of a type that represents the theorem. Interactive editors exist for both
types of systems. For Coq, the \texttt{Company-Coq} \cite{pitclaudel-2016-companycoq,aspinall-2000-general}
extension offers auto-completion, which recommends available tactics, hypotheses and local
definitions, but it does not filter them based on what is valid in a given context.
For Idris, the interactive editor \cite{brady-2015-idris} offers a range of commands that
transform the selected term by adding a case split, a missing case or by automatically searching
for a proof. In Idris, the system produces valid completions, but those cover only a small number
of situations.

Despite the different ways of working, an implementation of the choose-your-own-adventure
pattern for both types of systems would be similar. Based on the sub-goal that the user is
currently proving, the system would recommend a range of tactics that can be applied to the
sub-goal. In the case of Coq, the selected tactic would be added to the sequence. In Idris, the
selected tactic would be applied to transform the current term. The difference is in the mode of
embedding. A system for Coq would provide internal embedding in that the selected option is added
to the proof source code. (Much like selecting a completion when using type providers appends a
member access.) A system for Idris-like language would use the tactic to transform the term,
making it impossible to reconstruct the sequence of applied tactics as in the external mode of
embedding.

\begin{figure}
\[
\begin{array}{lcl}
\boxed{\Gamma \vdash \tau \Rightarrow e}\\[1em]
\dfrac{}{\Gamma,x:\tau \vdash \tau \Rightarrow x}~\ident{(syn-var)} &\quad&
\dfrac{}{\Gamma \vdash \tau \Rightarrow ?_\tau}~\ident{(syn-hole)}\\[2em]
\dfrac{\Gamma,x:\tau_1\vdash \tau_2 \Rightarrow e}{\Gamma \vdash \tau_1 \rightarrow \tau_2 \Rightarrow \lambda x.e}~\ident{(syn-lambda)}&&
\dfrac{\Gamma \vdash \tau_1\rightarrow\tau_2 \Rightarrow e_1 \quad \Gamma \vdash \tau_1 \Rightarrow e_2}{\Gamma \vdash \tau_2 \Rightarrow e_1~e_2}~\ident{(syn-app)}\\[2em]
\dfrac{b\in \{\ident{true},\ident{false}\} }{\Gamma \vdash \ident{bool} \Rightarrow b}~\ident{(syn-bool)} &\quad&
\dfrac{\Gamma \vdash \tau \Rightarrow e_1 \quad \Gamma \vdash \tau \Rightarrow e_2 \quad \Gamma \vdash \ident{bool} \Rightarrow e}{\Gamma \vdash \tau \Rightarrow \ident{if}~e~\ident{then}~e_1~\ident{else}~e_2}~\ident{(syn-cond)}\\
\end{array}
\]
\caption{Illustrative set of simple type-directed program synthesis rules}
\label{fig:synths}
\end{figure}

\subparagraph{Type-directed synthesis.}
Thanks to the equivalence between programs and proofs, techniques akin to tactic-based proof
construction have also emerged in work on type-directed program synthesis \cite{knoth-2023-synthesis}.
As illustrated in Figure~\ref{fig:synths}, the synthesis process can be described as a set of
rules of the form $\Gamma \vdash \tau \Rightarrow e$ that describe ways of synthesizing
expressions $e$ of a type $\tau$. Existing implementations of the mechanism typically aim to
automate program synthesis and use more precise type information, such as refinement types
\cite{polikarpova-2016-synthesis} and graded types \cite{hughes-2024-synthesis}, or include
examples \cite{osera-2015-synthesis}. However, the same rules could be used to guide an
interactive choose-your-own-adventure system. If the interaction was invoked to fill a typed
hole $?_\tau$ in a context $\Gamma$, the system could collect multiple $e$ such that
$\Gamma \vdash \tau \Rightarrow e$ and offer a choice of such options. Note that the definition
in Figure~\ref{fig:synths} synthesizes sub-expressions recursively, but a choose-your-own-adventure
system may always fill those with a typed hole using (\ident{syn-hole}).

\subparagraph{Formal model.}
From the perspective of user interaction, a proof assistant where a user interactively constructs
a term of a given type is very similar to an interactive tool for type-directed program synthesis.
The key difference being that theorem provers like Idris and Agda use rich dependent type theories.

For example, consider a system akin to Idris where the user aims to construct a term $e$ of type
$\tau$. The term may contain typed holes written as $?_\tau$ and a relation $\Gamma\vdash \tau\Rightarrow e$
provides ways of synthesizing terms of type $\tau$. We again write $E[?_\tau]$ for a
completion context containing a (typed) hole; we assume that the variables $\Gamma$ available in
the completion context of the hole can be obtained using $\ident{vars}(E[\_])$.

To model a choose-your-own-adventure interaction akin to Idris, the state of the system
would be the term $e$ itself, initially a typed hole. The $\choices$ operation synthesizes
possible completions using $\Rightarrow$ (restricted, e.g., to only generate terms of a certain
maximum size) and offers the resulting terms as possible completions. The identifiers $\iota$
could be based either on the tactic name (rule name) or show a preview of the resulting term.
The $\choices$ operation suggests ways to fill a hole in the term:
\[
\begin{array}{rcl}
\choices(E[?_\tau]) &=& \{\; \iota \mapsto e ~|~ \forall e\;.\; \ident{vars}(E[\_]) \vdash \tau \Rightarrow e_i \}\\
\select(e) &=& e
\end{array}
\]
The $\choices$ operation synthesizes possible terms of a type required by the hole. Since the
state $\sigma$ is a term, the $\select$ operation simply returns it. The definition models the
external embedding of the interaction, i.e., a system that behaves according to Idris. It
constructs the term, but does not record the completion choices. That said, a system akin to Coq
that constructs a sequence of tactics could be modelled too if the state was a sequence of
tactics and $\choices$ appended the available tactics as options to the end of the current sequence.

@@

ways of integrating automation with manual interaction
- confirm each step vs. auto and retrace

@@

\cite{verter-2024-mixed}
